\begin{appendices}
\section*{Use of Artificial Intelligence Tools}
\addcontentsline{toc}{section}{Use of Artificial Intelligence Tools}  

During the completion of this master's thesis, generative artificial intelligence tools were utilized in a limited and supportive capacity. This appendix provides a transparent account of how these tools were employed throughout the research and writing process.

\textbf{Ideation and Initial Development}

Generative AI was used as a brainstorming tool to explore initial concepts and generate starting points for various aspects of the implementation. These tools helped in considering different architectural approaches and identifying potential challenges in the early stages of the project. However, all final decisions regarding the technical approach, implementation strategy, and research direction were made independently through critical evaluation and analysis.

\textbf{Language and Writing Assistance}

AI tools were employed to enhance the fluency and readability of English text throughout this thesis. Given that English is not the author's native language, these tools served primarily as a translation and language refinement aid. The AI was used to:

\begin{itemize}
    \item Improve sentence structure and flow
    \item Enhance clarity of technical explanations  
    \item Ensure consistent academic writing style
    \item Correct grammatical errors and improve readability
\end{itemize}

It is important to emphasize that all conceptual content, technical insights, analysis, and conclusions presented in this thesis are entirely the author's original work. The AI tools were used exclusively for linguistic improvement and did not contribute to the intellectual substance of the research.

\textbf{Code Development and Error Detection}

Throughout the software development process, AI tools were utilized to identify potential errors and suggest improvements in code implementation. However, the author maintained full control over the development process by:

\begin{itemize}
    \item Critically reviewing all AI-generated suggestions
    \item Understanding and verifying the logic behind any proposed solutions
    \item Re-implementing all code sections independently rather than using direct copy-paste methods or allowing agentic operation of AI in the IDE or terminals
    \item Testing and validating all implementations thoroughly
\end{itemize}

All final code implementations, architectural decisions, and technical solutions represent the author's independent work and understanding of the problem domain.

\textbf{Academic Integrity Statement}

The use of AI tools as described above was conducted in accordance with academic integrity principles. The core research contribution, technical implementation, experimental design, analysis, and conclusions of this thesis remain entirely the work of the author. AI tools served solely as supportive instruments for language enhancement and development assistance, without compromising the originality or intellectual merit of the research.

This declaration is provided in the interest of transparency and adherence to evolving academic standards regarding the use of AI tools in scholarly work.



% \newpage
% \section*{Attachment B}
% \addcontentsline{toc}{section}{Attachment B}  

% Attachment description.

\end{appendices}