\chapter*{Acknowledgment}
\addcontentsline{toc}{chapter}{Acknowledgments}

I would like to express my sincere gratitude to all those who contributed to the successful completion of this master's thesis.

First and foremost, I extend my heartfelt thanks to my promotors, Prof. Dr. Bruno Volckaert and Dr. Merlijn Sebrechts, for placing their trust in me and providing the opportunity to participate in this research. Your confidence in my abilities has been invaluable throughout this journey.

Later in the year, Friedrich Vandenberghe, MSc in Industrial Engineering, joined our team as my predecessor on this thesis topic. Friedrich, your extensive experience with this research theme and your valuable contributions from the middle phase of my trajectory on have significantly enriched my work. Your insights and willingness to share your knowledge have been tremendously helpful.

I want to especially acknowledge Merlijn, Michiel, and Friedrich together again for always being available whenever I wanted to discuss results, share ideas, or seek advice. Your open-door policy and genuine interest in my progress have made this research journey both productive and enjoyable.

Finally, I would like to thank my girlfriend, friends, and parents for their unwavering support and genuine interest in my research. Your encouragement and curiosity about my work provided the extra motivation I needed to strive for excellent results. Your belief in me has been a constant source of strength throughout this challenging but rewarding experience.

This thesis would not have been possible without the collective support, guidance, and encouragement of all the individuals mentioned above. Thank you for making this journey memorable and successful.